---
title: "Student Funding and Graduation Outcomes"
author: "Scott Burghart"
date: "`r Sys.Date()`"
mainfont: Georgia
output: 
  bookdown::html_document2:
    df_print: kable
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = -2)
```

```{r LibraryLoad, include=FALSE}
# Load the libraries ####
library(readxl)
library(tidyverse)
library(lubridate)
library(ggpubr)
library(tools)
library(plotly) # for interactive plots
library(broom) # for r^2 values
UTColors5 <- c("#9cadb7", "#005f86", "#00a9b7", "#f8971f", "#a6cd57")
UTColors3 <- c("#005f86", "#f8971f", "#a6cd57")
```

```{r LoadAndClean, include=FALSE}
# Load and clean the data ####
support <- readxl::read_xlsx("FinancialSupport.xlsx") %>% 
  select(!c("College",
         "Degree Sought", # these data only contain PhD candidates
         "Total UT Grad Degrees", 
         "Doctoral Degree Semester", #this gets replaced by LSE
         "Total Reported Placements",
         "College")) %>% 
  rename("Serial" = "serial",
         "Cohort" = "Cohort Year",
         "MajorCode" = "Major Code",
         "LSE" = "Last Semester Enrolled",
         "TIP" = "ttd",
         "CandidacySem" = "Doctoral Candidacy Semester",
         "Status" = "Enrollment Status",
         "Supervisor1" = "Doctoral Supervisor 1",
         "Supervisor2" = "Doctoral Supervisor 2",
         "TotalSupport" = "Total Financial Support ($)",
         "AITA" = "AI/TA Appointments ($)",
         "GRA" = "GRA Appointments ($)",
         "Fellowship" = "Scholarship / Fellowship ($)",
         "OtherAC" = "Other Academic Appointments ($)",
         "Other" = "Other Support ($)") %>% 
  filter(!TIP == 0) %>% # filter out students who never enrolled in the  program or received no support
  mutate(AnnAITA = AITA / TIP,
         AnnGRA = GRA / TIP,
         AnnFllw = Fellowship / TIP,
         AnnOtherAC = OtherAC / TIP,
         AnnOther = Other / TIP,
         PercAITA = 100 * AITA / TotalSupport) %>%  # Create average annual amount of each type of support
  mutate(TotalSupport = TotalSupport / 1000,
         AITA = AITA / 1000,
         GRA = GRA / 1000,
         Fellowship = Fellowship / 1000,
         OtherAC = OtherAC / 1000,
         Other = Other / 1000) %>% # Change support units to 1000's of dollars
  mutate_at(vars(Cohort), factor) %>% 
  mutate(Supervisor1 = str_trim(Supervisor1, side = "both")) %>% # Strip whitespace from supervisor names
  mutate(Supervisor2 = str_trim(Supervisor2, side = "both")) %>% 
  # Change character strings in department and supervisor columns to title case
  mutate(Department = str_to_title(Department)) %>% 
  mutate(Supervisor1 = str_to_title(Supervisor1)) %>% 
  mutate(Supervisor2 = str_to_title(Supervisor2))
 

# Correct individual supervisor names
support$Supervisor1 <- sub("Blumberg, Andrew Justin", "Blumberg, Andrew J", support$Supervisor1)
support$Supervisor1 <- sub("Chen, Z. J", "Chen, Zengjian J", support$Supervisor1)
support$Supervisor1 <- sub("Contreras, Lydia Maria", "Contreras, Lydia M", support$Supervisor1)
support$Supervisor1 <- sub("Dickinson, Daniel James", "Dickinson, Daniel J", support$Supervisor1)
support$Supervisor1 <- sub("Downer, Michael Wayne", "Downer, Michael W", support$Supervisor1)
support$Supervisor1 <- sub("Freeland, Jeanne H", "Freeland-Graves, Jeanne H", support$Supervisor1)
support$Supervisor1 <- sub("Harris, R A", "Harris, R Adron", support$Supervisor1)
support$Supervisor1 <- sub("Winget, Don", "Winget, Donald E", support$Supervisor1)
sorted_Supervisor1 <- sort(support$Supervisor1)
unique(sorted_Supervisor1)

# Change units on support columns to 1000's of dollars
support$TotalSupport

# Change character vectors to title case
support$Department <- str_to_title(support$Department, locale = "en")

# Change character vectors to dates
support$StartDate <- mdy(support$StartDate)
support$EndDate <- mdy(support$EndDate)

# Transform Degree Outcome (Status) as a factor
support$Status <- as.factor(support$Status)

# Data containing only students who are no longer in the program
completed <- support %>% 
  filter(!Status == "DC" &
           !Status == "NC")
```


# Introduction

This report analyses the effect that the amount of TA funding has on the degree outcomes and the time to degree for graduate students in the College of Natural Sciences at UT Austin for cohorts 2012-2017. The below analyses involve only students who entered a graduate program seeking a PhD. Data come from the Graduate School Information System (GSIS). The Neuroscience program is not included because their data are not available for download via GSIS.

# Total Support

```{r fig.cap= "The relationship between the total amount of support received by students and their degree outcomes", echo=FALSE}
# Degree outcomes and distribution total funding ####
TotalSup <- ggplot(completed, aes(x = Status, y = TotalSupport, fill = Status)) +
  geom_violin() +
  scale_fill_manual(labels = c("DM - Master's", "DT - PhD", "NX - No degree"), values = UTColors5) +
  theme_bw() +
  theme(legend.position = "top") +
  ylab("Total Support ($1,000's)") +
  geom_point(position = position_jitter(seed = 1, width = 0.2))
TotalSup
```


```{r Normal_Test_Total_Support, include=FALSE}
# Test for normality of total support data #####
# Create a Q-Q plot
qqnorm(completed$TotalSupport)
qqline(completed$TotalSupport)
shapiro.test(completed$TotalSupport)

# Data are not normally distributed - Kruskall-Wallis test for differences #####
kruskal.test(TotalSupport ~ Status, data = completed)

```

```{r Mean_Total_Support, include=FALSE}
TotalSummary <- support %>% 
  group_by(Status) %>% 
  summarize(MeanTotal = mean(TotalSupport))
```


Unsurprisingly, significantly more money is ultimately spent on students who finish their degrees. For students who left with a PhD, an average of &dollar;296,297 was invested, compared to &dollar;162,574 for students departing with a master's and &dollar;92,435 for students leaving with no degree. As an aside, though not included in the figure, due to increasing costs of student support, it was calculated taht students currently enrolled in PhD programs have already received an average of &dollar;322,937.

# Form of Support

Of more interest for this report is the form student support takes and its effects on degree outcomes and time to degree (TTD). Specifically, do students who receive more support in the form of AI or TA (AITA) assignments take longer to finish or are they at a greater risk of departing the program without completing their PhD? This is commplicated by the fact that the number of semesters a student spends as an AITA is not readily available. The dollar amount of support as an AITA can be used as a proxy for the number of semesters the student was employed in an instructional role but this is a less than optimal measure of AITA semesters since the cohorts examined span a time of aggressive stipend increases within the college. As such, AITA semesters towards the end of the period examined will be weighted more heavily. Additionally, AI assignments pay 6% more than TA assignments and the source data makes no distinction between the two. To partially control for these issues, the percentage of total support as an AITA was used.

## Degree Outcomes

```{r fig.cap = "The relationship between percent of support as an AITA and degree outcomes", echo=FALSE, warning=FALSE}
AITASupPerc <- ggplot(completed, aes(x = Status, y = PercAITA, fill = Status)) +
  geom_violin() +
  scale_fill_manual(labels = c("DM - Master's", "DT - PhD", "NX - No degree"), values = UTColors5) +
  theme_bw() +
  theme(legend.position = "top") +
  ylab("Percent of support as AITA") +
  geom_point(position = position_jitter(seed = 1, width = 0.2))
AITASupPerc
```

```{r Normal_Test_PercAITA, include=FALSE}
# Test for normality of total support data #####
# Create a Q-Q plot
qqnorm(completed$PercAITA)
qqline(completed$PercAITA)
shapiro.test(completed$PercAITA)

# Data are not normally distributed - Kruskall-Wallis test for differences #####
kruskal.test(PercAITA ~ Status, data = completed)

```

```{r Mean_Percent_AITA, include=FALSE}
PercAITASummary <- completed %>% 
  filter(!is.na(PercAITA)) %>% 
  group_by(Status) %>% 
  summarize(MeanPercAITA = mean(PercAITA))
```


<br/>

The percentage of total support attributed to AITA work assignments was different between the degree outcome groups (Kruskal-Wallis p < 0.05). On average students who finished with a PhD received 19.9% of their support as an AITA while those who left without a degree averaged 24.5% and those who finished with a master's received 32.6%.

It is important to note that students entering two of the largest programs (Chemistry and Physics), almost invariably TA their first year. This will increase the average annual amount of AITA support for students who stay less than five years and thus will, thus, usually appear in the DM or NX category. To examine if these two programs had an undue effect, they were removed and the analysis was repeated.

```{r fig.cap= "The relationship between percentage support as an AITA and degree outcomes with Physics and Chemistry removed", echo=FALSE, warning=FALSE}
# Remove Physics and Chemistry from the analysis
completed2 <- completed %>% 
  filter(!(Department %in% c("Physics", "Chemistry")))

AITASup2Perc <- ggplot(completed2, aes(x = Status, y = PercAITA, fill = Status)) +
  geom_violin() +
  scale_fill_manual(labels = c("DM - Master's", "DT - PhD", "NX - No degree"), values = UTColors5) +
  theme_bw() +
  theme(legend.position = "top") +
  ylab("Percentage AITA Support") +
  geom_point(position = position_jitter(seed = 1, width = 0.2))
AITASup2Perc
```


```{r Mean_AITA2_Percent, include=FALSE}
AITASummary2Perc <- completed2 %>% 
  filter(!is.na(PercAITA)) %>% 
  group_by(Status) %>% 
  summarize(MeanPercAITA = mean(PercAITA))

```

```{r Normal_Test_TASupport2_Percent, include=FALSE}
# Test for normality of total support data #####
# Create a Q-Q plot
qqnorm(completed2$PercAITA)
qqline(completed2$PercAITA)
shapiro.test(completed2$PercAITA)

# Data are not normally distributed - Kruskal-Wallis test for differences #####
kruskal.test(PercAITA ~ Status, data = completed2)
```

<br/>

With Chemistry and Physics students removed from the analysis, the percentage of funding from being an AITA was virtually identical between students who finished with a PhD and those who left with no degree at 17.7% and 16.0%, respectively. The average percentage of AITA support for students that left with a master's, however, was almost twice as much at 31.1%

<br/>

## Time to Degree

```{r fig.cap="Relationship between the percentage of AITA support and TTD", echo=FALSE, message=FALSE, warning=FALSE}
TAeffectPerc <- ggplot(completed, aes(x = PercAITA, y = TIP, color = Status)) +
  scale_color_manual(values = UTColors3) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw() +
  ylab("Time in Program (years)") +
  xlab("Percent Support as AITA")
TAeffectPerc
```

<br/>

``` {r fig.cap="Relationship between percentage of AITA funding and TTD between different programs", echo=FALSE}
# Stats by Supervisor1 and department ####
support_by_PI_perc <- support %>% 
  group_by(Department, Supervisor1) %>% 
  summarize(mean(PercAITA))

time_by_PI <- support %>% 
  group_by(Department, Supervisor1) %>% 
  summarize(mean(TIP))

PercAITA_by_PI <- support %>% 
  group_by(Department, Supervisor1) %>% 
  summarize(mean(PercAITA))

time_by_dept <- support %>% 
  filter(Status == "DT") %>% 
  group_by(Department) %>% 
  summarize(mean(TIP))

PercAITA_by_dept <- support %>% 
  filter(Status == "DT") %>% 
  group_by(Department) %>% 
  summarize(mean(PercAITA))

time_funding_by_dept_perc <- left_join(time_by_dept, PercAITA_by_dept, by = "Department") %>% 
  rename("AvgTIP" = "mean(TIP)") %>% 
  rename("AvgPercAITA" = "mean(PercAITA)")

# Time in program in each department by the average annual amount of TA support ####
dept_time_by_TA_perc <- ggplot(time_funding_by_dept_perc, aes(x = AvgPercAITA, y = AvgTIP, color = Department)) +
  geom_point(size = 5) +
  scale_color_brewer(palette = "Paired") +
  theme_bw() +
  theme(legend.position = "top") +
  ylab("Average Time to Graduation (years)") +
  xlab("Average Percentage Support as AITA")
dept_time_by_TA_perc

```

<br/><br/>

```{r fig.cap="Relationship between average annual AITA funding and TTD of PhD students in different programs", echo=FALSE, message=FALSE, warning=FALSE}
# Function to get R-squared value
get_r_squared <- function(model) {
  rsq <- summary(model)$r.squared
  return(rsq)
}

PI_wn_pgm_perc <- support %>% 
  filter(Status == "DT") %>% 
  group_by(Department, Supervisor1) %>% 
  summarise(
    "PercTAsup" = mean(PercAITA),
    "TIP" = mean(TIP)
  )

# Create a summary data frame with R-squared values
r_squared_df <- PI_wn_pgm_perc %>% 
  group_by(Department) %>% 
  summarise(R_squared = get_r_squared(lm(TIP ~ PercTAsup)))

# Merge R-squared values back to the main data frame
PI_wn_pgm_perc <- merge(PI_wn_pgm_perc, r_squared_df, by = "Department")

# Create the plot
PIsupportPerc <- ggplot(PI_wn_pgm_perc, aes(x = PercTAsup, y = TIP, color = Department)) +
  geom_point(size = 1.5) +
  stat_smooth(method = "lm", se = FALSE, fullrange = TRUE, color = "#005f86") +
  scale_color_brewer(palette = "Paired") +
  theme_bw() +
  theme(legend.position = "none") +
  xlab("Average percentage support as AITA") +
  ylab("Average time to degree (years)") +
  facet_wrap(~ Department, scales = "free") +  # Add separate plots for each Department
  
  # Use geom_text to add R-squared values
  geom_text(
    aes(
      x = Inf, y = -Inf, 
      label = sprintf("R^2 = %.2f", R_squared)
    ),
    data = r_squared_df, # Use the summary data frame
    hjust = 1.1, vjust = -0.2,
    color = "#005f86",
    size = 3,
    show.legend = FALSE
  )

PIsupportPerc
```

<br/><br/>

Across the college, time to degree (TTD) did not show a positive relationship to the percentage of support as an AITA.

On a departmental basis, there is a gap splitting the programs up into two different groups. Marine Science, ILS (Molec Biosciences), Astronomy, and Computer Science all rely on AITA assignments to provide less than 15% of student funding. Chemistry, Integrative Biology (EEB and PB), Physics, Mathematics, SDS, HDFS, and Nutritional Sciences all rely on AITA assignments to provide more than 20% of their student funding. Nutritional Sciences relies on AITA assignments by far the most, using them to provide 36% of student funding. Between departments, though, for students completing their PhD, the average percentage of total support as an AITA showed a relationship with time to degree.

In Figure 3.5, each point represents an individual PI within each program. Within each department, the average TTD versus percentage support as an AITA did have some positive slopes, but R-squared values were very low. The highest values occurred in Integrative Biology and Computer Science.